*Symbiodinium* communities in reef corals of St. John, U.S.V.I. (2012)

### 2016-08-24 analysis/STJ2012_analysis.Rmd

# Filter samples by minimum count
# Set threshold number of reads
sn <- 200
# Remove samples with fewer reads than threshold
phy97.f <- prune_samples(sample_sums(phy97.f)>=sn, phy97.f)
phy97bs.f <- prune_samples(sample_sums(phy97bs.f)>=sn, phy97bs.f)
phy97bsp.f <- prune_samples(sample_sums(phy97bsp.f)>=sn, phy97bsp.f)
phy100.f <- prune_samples(sample_sums(phy100.f)>=sn, phy100.f)
```
* Minimum count to retain OTU: `r n`
* Minimum count to retain sample: `r sn`

@@

# Clade overview
Here the composition of each sample is plotted as a horizontal bar, sorted by species and shore. Each segment of the bar represents an individual OTU and is colored by clade. This plot is a nice overview of the whole dataset but only provides coarse information at the clade level.
```{r cladecomp97bs, cache=T, fig.height=6, fig.width=4}
par(mfrow=c(1,1), mar=c(2, 1.5, 2, 5), lwd=0.1, cex=0.7)
# Plot composition of 97% within-sample OTUs colored by clade
composition(phy97bs.f.p, col=taxcolors[factor(data.frame(tax_table(phy97bs.f.p))[order(data.frame(tax_table(phy97bs.f.p))$Subtype), ]$Clade, levels=c("A","B","C","D","F","G"))], legend=T)
# Plot composition of 97% within-sample OTUs colored by OTU
#composition(phy97bs.f.p, col=rainbow(ntaxa(phy97bs.f.p)), legend=F)

@@

# Network analyses
## "Resident" symbionts
#### (>1% relative abundance)
In this network, each coral species is collapsed into a single node (white squares), and connected to every *Symbiodinium* OTU (colored circles) that comprises **at least 1% relative abundance** within at least one individual of that species. Thus, very low abundance OTUs are not represented. The thickness of connections between coral species and OTUs is scaled by the relative proportion of samples of that species in which that OTU was present at ≥1% relative abundance (i.e., the frequency of presence at ≥1%). The size of *Symbiodinium* OTU nodes in the network is scaled by the number of species in which they occur, and colored according to clade.
```{r networks}
# Get network with one node for each species.
otudf <- data.frame(t(otu_table(phy97bs.f.p)), check.names=F)
# Percent of samples of species that contain OTU at more than 1% relative abundance
sp.ag <- aggregate(otudf, by=list(Species=data.frame(sample_data(phy97bs.f.p))$Species), 
                   FUN=function(x) length(x[x>0.01])/length(x))
rownames(sp.ag) <- sp.ag$Species
sp.ag <- sp.ag[, -1]
edges <- setNames(melt(as.matrix(sp.ag)), nm=c("source", "target", "weight"))  # Melt data
edges <- droplevels(edges[edges$weight>0, ])
# Create "nodes" table from sample and tax data
sdf <- data.frame(id=levels(data.frame(sample_data(phy97bs.f.p))$Species))
tdf <- data.frame(tax_table(phy97bs.f.p))[rownames(data.frame(tax_table(phy97bs.f.p))) %in% edges$target, ]
tdf$id <- rownames(tdf)
nodes <- merge(sdf, tdf, all=T)
nodes$type <- ifelse(is.na(nodes$Clade), 1, 2)
# Create network
net <- graph_from_data_frame(d=edges, vertices=nodes, directed=F)

@@

## Dominant symbionts
#### (>50% relative abundance)
```{r dom_networks}
# Get network with one node for each species.
otudf <- data.frame(t(otu_table(phy97bs.f.p)), check.names=F)
# Percent of samples of species that contain OTU at more than 1% relative abundance
sp.ag <- aggregate(otudf, by=list(Species=data.frame(sample_data(phy97bs.f.p))$Species), 
                   FUN=function(x) length(x[x>0.5])/length(x))
rownames(sp.ag) <- sp.ag$Species
sp.ag <- sp.ag[, -1]
edges <- setNames(melt(as.matrix(sp.ag)), nm=c("source", "target", "weight"))  # Melt data
edges <- droplevels(edges[edges$weight>0, ])
# Create "nodes" table from sample and tax data
sdf <- data.frame(id=levels(data.frame(sample_data(phy97bs.f.p))$Species))
tdf <- data.frame(tax_table(phy97bs.f.p))[rownames(data.frame(tax_table(phy97bs.f.p))) %in% edges$target, ]
tdf$id <- rownames(tdf)
nodes <- merge(sdf, tdf, all=T)
nodes$type <- ifelse(is.na(nodes$Clade), 1, 2)
# Create network
net <- graph_from_data_frame(d=edges, vertices=nodes, directed=F)

@@

## Background symbionts
#### (<1% relative abundance)
```{r bg_net}
# Get network with one node for each species.
otudf <- data.frame(t(otu_table(phy97bs.f.p)), check.names=F)
# Percent of samples of species that contain OTU at less than 1% relative abundance (but more than 0%)
# Only record if bkgd symbiont was detected in more than one individual
sp.ag <- aggregate(otudf, by=list(Species=data.frame(sample_data(phy97bs.f.p))$Species), 
                   FUN=function(x) ifelse(length(x[x<0.01 & x>0])>1, # if more than 1 individual has bkgd symbiont
                                          length(x[x<0.01 & x>0])/length(x), 0))
sp.ag <- aggregate(otudf, by=list(Species=data.frame(sample_data(phy97bs.f.p))$Species), 
                   FUN=function(x) length(x[x<0.01 & x>0])/length(x))
sp.ag <- aggregate(otudf, by=list(Species=data.frame(sample_data(phy97bs.f.p))$Species), 
                   FUN=function(x) any(x>0 & x<0.01))
rownames(sp.ag) <- sp.ag$Species
sp.ag <- sp.ag[, -1]
edges <- setNames(melt(as.matrix(sp.ag)), nm=c("source", "target", "weight"))  # Melt data
edges <- droplevels(edges[edges$weight>0, ])
# Create "nodes" table from sample and tax data
sdf <- data.frame(id=levels(data.frame(sample_data(phy97bs.f.p))$Species))
tdf <- data.frame(tax_table(phy97bs.f.p))[rownames(data.frame(tax_table(phy97bs.f.p))) %in% edges$target, ]
tdf$id <- rownames(tdf)
nodes <- merge(sdf, tdf, all=T)
nodes$type <- ifelse(is.na(nodes$Clade), 1, 2)
# Create network
net <- graph_from_data_frame(d=edges, vertices=nodes, directed=F)

# Best one for now ---------
V(net)$shape <- c("square", "circle")[factor(V(net)$type)]
V(net)$size <- ifelse(is.na(V(net)$Clade), 15, 5*degree(net)) #5*sqrt(degree(net))
V(net)$label <- ifelse(is.na(V(net)$Clade), names(V(net)),  
                       unlist(lapply(strsplit(V(net)$Subtype, split="_"), "[", 1)))
V(net)$label <- str_replace_all(V(net)$label, c("alcicornis"="M.alc.", "annularis"="O.ann",
                                                "astreoides"="P.ast.", "cavernosa"="M.cav.",
                                                "cylindrus"="D.cyl.", "fragum"="F.fra.", 
                                                "furcata"="P.fur.", "radians"="S.rad.", 
                                                "siderea"="S.sid.", "strigosa"="D.str."))
V(net)$color <- ifelse(is.na(V(net)$Clade), "white",
                       taxcolors[factor(V(net)$Clade, levels=c("A","B","C","D","F","G"))])
E(net)$color <- "gray60"
E(net)$width <- 1
set.seed(12347)
set.seed(12364)
set.seed(12374)
l <- layout.fruchterman.reingold(net)
l <- norm_coords(l, ymin=-1, ymax=1, xmin=-1, xmax=1)
par(mar=c(0,0,0,0))
#plot(net, rescale=F, layout=l*1, edge.curved=0.1, vertex.label.cex=V(net)$size/20,
#     vertex.label.color="black")
#plot(net, layout=layout_nicely, edge.curved=0.1, vertex.label.cex=V(net)$size/20,
#     vertex.label.color="black")
#plot(net, layout=layout_as_tree(net, circular=T))
#plot(net, layout=layout_with_dh)

net2 <- delete.vertices(net, degree(net)<=3)
V(net2)$size <- ifelse(is.na(V(net2)$Clade), 15, 4*degree(net2))
set.seed(42381)
l2 <- layout.fruchterman.reingold(net2)
l2 <- norm_coords(l2, ymin=-1, ymax=1, xmin=-1, xmax=1)
plot(net2, rescale=F, layout=l2*1, edge.curved=0.1, vertex.label.cex=V(net2)$size/20,
     vertex.label.color="black")
#plot(net2, layout=layout_as_tree(net2, circular=T), edge.curved=0.1)
#plot(net2, layout=layout_as_star(net2, center = 18))
#plot(net2, layout=layout_in_circle(net2))
#plot(net2, layout=layout_with_kk)
#plot(net2, layout=layout_as_tree(net2, root=16, circular=T))
#plot(net2, layout=layout_with_dh)
#?layout_as_tree
#V(net2)

#cor<-induced.subgraph(net2,V(net2)$type==1)
#sym<-induced.subgraph(net2,V(net2)$type==2)
#polar.layout <- function(radii, angles) {
#    cbind(radii*cos(angles), -radii*sin(angles))        
#}
#cor <- which(V(net2)$type == 1)
#sym <- which(V(net2)$type == 2)
#radii <- ifelse(V(net2)$type == 1, 2, 1)
#angles <- rep.int(0, vcount(net2))
#angles[sym] <- (1:length(sym)-1) * 2 * pi / length(sym)
#angles[cor] <- (1:length(cor)-1) * 2 * pi / length(cor)
#layout <- polar.layout(radii, angles)
#plot(net2, layout=layout)
```

>>>>

# Filter samples by minimum count
# Set threshold number of reads
sn <- 200
# Remove samples with fewer reads than threshold
phy97.f <- prune_samples(sample_sums(phy97.f)>=sn, phy97.f)
phy97bs.f <- prune_samples(sample_sums(phy97bs.f)>=sn, phy97bs.f)
phy97bsp.f <- prune_samples(sample_sums(phy97bsp.f)>=sn, phy97bsp.f)
phy100.f <- prune_samples(sample_sums(phy100.f)>=sn, phy100.f)

# Filter OTUs by minimum count again in case any dropped below threshold after filtering samples
# Identify OTUs below threshold count
taxa97 <- taxa_sums(phy97.f)[which(taxa_sums(phy97.f) >= n)]
taxa97bs <- taxa_sums(phy97bs.f)[which(taxa_sums(phy97bs.f) >= n)]
taxa97bsp <- taxa_sums(phy97bsp.f)[which(taxa_sums(phy97bsp.f) >= n)] 
taxa100 <- taxa_sums(phy100.f)[which(taxa_sums(phy100.f) >= n)]
# Remove taxa below threshold count
phy97.f <- prune_taxa(names(taxa97), phy97.f)
phy97bs.f <- prune_taxa(names(taxa97bs), phy97bs.f)
phy97bsp.f <- prune_taxa(names(taxa97bsp), phy97bsp.f)
phy100.f <- prune_taxa(names(taxa100), phy100.f)
```
* Minimum count to retain OTU: `r n`
* Minimum count to retain sample: `r sn`

@@

# Clade overview
Here the composition of each sample is plotted as a horizontal bar, sorted by species and shore. Each segment of the bar represents an individual OTU and is colored by clade. This plot is a nice overview of the whole dataset but only provides coarse information at the clade level.
```{r cladecomp97bs, cache=T, fig.height=6, fig.width=4}
cladeAbund <- aggregate(data.frame(RelAbund=rowSums(otu_table(phy97bs.f.p))),
                        by=list(Clade=data.frame(tax_table(phy97bs.f.p))$Clade), FUN=sum)
sum(cladeAbund$RelAbund)
cladeAbund$Prop <- prop.table(cladeAbund$RelAbund)
pie(cladeAbund$Prop, col=taxcolors)
dotchart(cladeAbund$Prop)
bars <- barplot(cladeAbund$Prop*100, col=taxcolors, space=0,
                names.arg=cladeAbund$Clade, xlab=expression(paste(italic('Symbiodinium'), " Clade")),
                ylab="Relative abundance (%)")
text(bars, cladeAbund$Prop*100+2, labels=round(cladeAbund$Prop*100, 1), xpd=T)

cladeAbund$Notus <- table(data.frame(tax_table(phy97bs.f.p))$Clade)
cladeAbund

par(mfrow=c(1,1), mar=c(2, 1.5, 2, 5), lwd=0.1, cex=0.7)
# Plot composition of 97% within-sample OTUs colored by clade
composition(phy97bs.f.p, col=taxcolors[factor(data.frame(tax_table(phy97bs.f.p))[order(data.frame(tax_table(phy97bs.f.p))$Subtype), ]$Clade, levels=c("A","B","C","D","F","G"))], legend=T)
# Plot composition of 97% within-sample OTUs colored by OTU
#composition(phy97bs.f.p, col=rainbow(ntaxa(phy97bs.f.p)), legend=F)

@@

# Network analyses
## "Resident" symbionts
#### (>1% relative abundance)
In this network, each coral species is collapsed into a single node (white squares), and connected to every *Symbiodinium* OTU (colored circles) that comprises **at least 1% relative abundance** within at least one individual of that species. Thus, very low abundance OTUs are not represented. The thickness of connections between coral species and OTUs is scaled by the relative proportion of samples of that species in which that OTU was present at ≥1% relative abundance (i.e., the frequency of presence at ≥1%). The size of *Symbiodinium* OTU nodes in the network is scaled by the number of species in which they occur, and colored according to clade.
```{r networks, fig.width=5, fig.height=5, fig.align="center"}
# Get network with one node for each species.
otudf <- data.frame(t(otu_table(phy97bs.f.p)), check.names=F)
# Percent of samples of species that contain OTU at more than 1% relative abundance
sp.ag <- aggregate(otudf, by=list(Species=data.frame(sample_data(phy97bs.f.p))$Species), 
                   FUN=function(x) length(x[x>0.01])/length(x))
rownames(sp.ag) <- sp.ag$Species
sp.ag <- sp.ag[, -1]
edges <- setNames(melt(as.matrix(sp.ag)), nm=c("source", "target", "weight"))  # Melt data
edges <- droplevels(edges[edges$weight>0, ])
# Create "nodes" table from sample and tax data
sdf <- data.frame(id=levels(data.frame(sample_data(phy97bs.f.p))$Species))
tdf <- data.frame(tax_table(phy97bs.f.p))[rownames(data.frame(tax_table(phy97bs.f.p))) %in% edges$target, ]
tdf$id <- rownames(tdf)
nodes <- merge(sdf, tdf, all=T)
nodes$type <- ifelse(is.na(nodes$Clade), 1, 2)
# Create network
net <- graph_from_data_frame(d=edges, vertices=nodes, directed=F)

@@

## Dominant symbionts
#### (>50% relative abundance)
```{r dom_networks, fig.width=5, fig.height=5, fig.align="center"}
# Get network with one node for each species.
otudf <- data.frame(t(otu_table(phy97bs.f.p)), check.names=F)
# Percent of samples of species that contain OTU at more than 50% relative abundance
sp.ag <- aggregate(otudf, by=list(Species=data.frame(sample_data(phy97bs.f.p))$Species), 
                   FUN=function(x) length(x[x>0.5])/length(x))
rownames(sp.ag) <- sp.ag$Species
sp.ag <- sp.ag[, -1]
edges <- setNames(melt(as.matrix(sp.ag)), nm=c("source", "target", "weight"))  # Melt data
edges <- droplevels(edges[edges$weight>0, ])
# Create "nodes" table from sample and tax data
sdf <- data.frame(id=levels(data.frame(sample_data(phy97bs.f.p))$Species))
tdf <- data.frame(tax_table(phy97bs.f.p))[rownames(data.frame(tax_table(phy97bs.f.p))) %in% edges$target, ]
tdf$id <- rownames(tdf)
nodes <- merge(sdf, tdf, all=T)
nodes$type <- ifelse(is.na(nodes$Clade), 1, 2)
# Create network
net <- graph_from_data_frame(d=edges, vertices=nodes, directed=F)

@@

## Background symbionts
#### (<1% relative abundance)
```{r bk.net, fig.width=5, fig.height=5, fig.align="center"}
# filter out OTUs that only occurred in one sample
phy97bs.f.p2 <- filter_taxa(phy97bs.f.p, function(x) sum(x>0) > 1, prune=TRUE)
# Get network with one node for each species.
otudf <- data.frame(t(otu_table(phy97bs.f.p2)), check.names=F)
# Percent of samples of species that contain OTU at less than 1% relative abundance (but more than 0%)

sp.ag <- aggregate(otudf, by=list(Species=data.frame(sample_data(phy97bs.f.p2))$Species), 
                   FUN=function(x) any(x>0 & x<0.01))
rownames(sp.ag) <- sp.ag$Species
sp.ag <- sp.ag[, -1]
edges <- setNames(melt(as.matrix(sp.ag)), nm=c("source", "target", "weight"))  # Melt data
edges <- droplevels(edges[edges$weight>0, ])
# Create "nodes" table from sample and tax data
sdf <- data.frame(id=levels(data.frame(sample_data(phy97bs.f.p2))$Species))
tdf <- data.frame(tax_table(phy97bs.f.p2))[rownames(data.frame(tax_table(phy97bs.f.p2))) %in% edges$target, ]
tdf$id <- rownames(tdf)
nodes <- merge(sdf, tdf, all=T)
nodes$type <- ifelse(is.na(nodes$Clade), 1, 2)
# Create network
net <- graph_from_data_frame(d=edges, vertices=nodes, directed=F)

# Edit vertex attributes
V(net)$shape <- c("square", "circle")[factor(V(net)$type)]
V(net)$size <- ifelse(is.na(V(net)$Clade), 15, 5*degree(net)) #5*sqrt(degree(net))
V(net)$label <- ifelse(is.na(V(net)$Clade), names(V(net)),  
                       unlist(lapply(strsplit(V(net)$Subtype, split="_"), "[", 1)))
V(net)$label <- str_replace_all(V(net)$label, c("alcicornis"="M.alc.", "annularis"="O.ann",
                                                "astreoides"="P.ast.", "cavernosa"="M.cav.",
                                                "cylindrus"="D.cyl.", "fragum"="F.fra.", 
                                                "furcata"="P.fur.", "radians"="S.rad.", 
                                                "siderea"="S.sid.", "strigosa"="D.str."))
V(net)$color <- ifelse(is.na(V(net)$Clade), "white",
                       taxcolors[factor(V(net)$Clade, levels=c("A","B","C","D","F","G"))])
E(net)$color <- "gray60"
E(net)$width <- E(net)$weight * 2

# Merge D nodes (assuming they are all D1a/trenchii)
# Get original node map (sequence starting at one to the total number of nodes)
nodemap <- seq(1:length(V(net)))
# Make a copy of the nodemap that will be modified
nodemapnew <- nodemap
# Find all nodes beginning with D, and give them the same node ID number
nodemapnew[grep("^D", V(net)$Subtype)] <- min(grep("^D", V(net)$Subtype))
# Replace the now missing values in the node map so node IDs remain sequential
while (max(setdiff(nodemap, nodemapnew)) < max(nodemap)) {
  nodemapnew[which(nodemapnew==max(setdiff(nodemap, nodemapnew))+1)] <- min(setdiff(nodemap, nodemapnew))
}
# Contract the network using the new nodemap (nodes with same node ID are merged)
net2 <- contract(net, nodemapnew, vertex.attr.comb=list(size="sum", "last"))
# Simplify network so that multiple edges are combined
net2 <- simplify(net2, remove.multiple=T, edge.attr.comb=list(weight="mean", width="mean", "ignore"))

# remove symbionts that are only connected to one or two coral species
net3 <- delete.vertices(net2, degree(net2)<=2) # & V(net2)$type==2)
V(net3)$size <- ifelse(is.na(V(net3)$Clade), 15, 1.5*degree(net3)^1.5)
set.seed(42384)
l3 <- layout.fruchterman.reingold(net3, weights=E(net3)$weight)
l3 <- norm_coords(l3, ymin=-1, ymax=1, xmin=-1, xmax=1)
par(mar=c(0,0,0,0))
plot(net3, rescale=F, layout=l3*1, edge.curved=0.1, vertex.label.cex=V(net3)$size/20,
     vertex.label.color="black")
points(0.9,0.9, pch=0, cex=4)
text(0.9,0.9, "O.ann.", cex=0.5)
